A Causal Model of \hbox{DBMS} Suboptimality\\

### 2017-03-14 subopt-tods16/paper.tex
SubOpt paper updates

%Metadata Information
\acmVolume{1}
\acmNumber{1}
\acmArticle{1}
\acmYear{2016}
\acmMonth{2}

\usepackage{url}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{graphics}
% \usepackage[total={6.5in,9in}, top=1.25in,left=0.9in]{geometry}
%\usepackage[text={6.5in,9in},centerpage,includefoot]{geometry}
% \usepackage{setspace}
\usepackage{color}
\usepackage{url}
\usepackage{balance}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{subfigure}

@@

\begin{abstract}
The query optimization phase within a \hbox{DBMS} ostensibly finds the fastest query
execution plan from a potentially large set of enumerated plans, all of
which correctly compute the specified query. Occasionally the optimizer
selects the wrong plan, for a variety of reasons. {\em Suboptimality} is
indicated by the existence of a query plan that performs more efficiently
than the DBMS's chosen plan, for the same query. From an engineering
perspective, it is of critical importance to understand the prevalence
of suboptimality and its causal factors.
We propose a novel structural causal model to explicate the
relationship between various factors in query optimization and
suboptimality. Our model associates suboptimality with the factors of
complexity of the query and optimizer and concomitant unanticipated
interactions among the components of the optimizer. This model induces a number of specific
hypothesis that were subsequently tested on multiple \hbox{DBMSes}. We
observe that \todo{Rick: check these}suboptimality prevalence
correlates positively with the number of operators available in the \hbox{DBMS}
(one proxy for optimizer complexity) and with the number of plans generated
by the optimizer (another proxy for optimizer complexity) and with
the number of correlation names (a measure of query complexity), providing
empirical support for this model as well as implications for fundamental
improvements of these optimizers. The results of this study also uncovered evidence of a potential fundamental
limitation of cost-based optimization.\todo{Rick:be more specific! Couple
  with contributions and conclusions}
This paper thus provides
a new methodology to study mature query optimizers, that of empirical
generalization, proposes a novel causal model for query suboptimality, and
tests several hypotheses deriving from this causal model.
\end{abstract}

@@

\item We test the six quite specific hypotheses that arise from that
  model across a wide range of DBMSes, queries, and data.

\item We show through correlational and regression analysis that four
  hypothesized interactions are strongly supported, one is partially supported, and
  an interaction hypothesized to go one direction went quite weakly the
  other direction, lending credibility to our particular model.

@@

\begin{figure}[h!]\centering
%\todo{Young: were is the figure!}
\includegraphics[width=0.60\textwidth]{figures/org_ncpq.pdf}
\caption{Cumulative percentage of queries \hbox{exhibiting} the \hbox{indicated}
  number of plan changes\label{fig:planchanges}}
\end{figure}

@@

This is why we needed a new methodology. We want to understand
cost-based optimization deeply. This means that we need to go beyond the
examination of a single query optimizer, as is done in almost every paper
over the forty-year history of query optimization research, to study multiple instances of
that optimization architecture, in an effort to achieve generalizable
results.

@@

Early investigation of plan suboptimality resulted in approaches such as
dynamic query-reoptimization~\cite{Avnur,kabra98}, which exploit more
accurate runtime statistics that appear while a query is being executed, to
steer in-flight plan reoptimization. The very presence of such a radical
change to the normal optimize-execute sequence indicates that plan
suboptimality was of interest to some researchers.

@@

\vspace{0.6em}
We don't feel that skewed data (in the ``data complexity''
construct) will impact plan space
complexity, but rather that it will negatively impact accuracy of plan cost
estimation, and thus increase suboptimality.

@@

\vspace{0.6em}
The schema complexity construct consists of the variable ``Presence of
secondary indexes.'' Such indexes provide opportunities for the optimizer to
consider more candidate plans that use these indexes, due to the additional
query evaluation operators now applicable. Those additional plans enable the optimizer to
possibly do a better job, while also adding complexity to the optimization
process. We hypothesize that this factor has a more complex role in the
model, serving as a {\em moderator} of two interactions previously
introduced. We hypothesize that the overall effect of the secondary indexes
is to increase the strength of the interaction between query
complexity to plan space complexity. (Note that as the presence of primary
key is required for secondary indexes, we don't include that former
independent variable here.)

@@

We define the skew quite simply: ``the reciprocal of the number of distinct
values,'' so $0 < \hbox{\em skew} \leq 1$.
For 2M distinct values, the skew would be $\frac{1}{2M}= 0.000005$, which is practically 0. 
For exactly one distinct value, the skew would be 1.0. For two distinct values, 
the skew would be 0.5. For ten distinct values, the skew would be 0.1.

@@

Let's first illustrate with a skew factor of 1.0. 
This translates to exactly 1 value for the entire table, or 2M spans.
To shorten, we remove 10K spans, equal to 10K values. 
This removal keeps the skew at 1.0, due to the unique values in the
remaining spans.
At the end, 10K spans, each with a single value, remain.

@@

Let's see an example at the other end of the spectrum, of a skew factor of
1/2M.  This skew factor, which is almost close to 0, gets translated to 2M
distinct values.  So we generate a single span of 2M values.  To shorten the
table, it makes no sense to remove this single span in its entirety. But we
can remove 10K values from this single span.  Note that this changes the
skew to 1/1.99M, then eventually to 1/10K, which is still very close to zero
(the skew has changed from .0000005 to .00001).

We use five skewness values: 0.000005 (which we term only slightly
misleadingly as ``no skew''), 0.001, 0.1, 0.5, and 1.0,
for this independent variable.

@@

To classify each operator as continuous or discontinuous, we use the data
from the Exhaustive Experiment. We
used a small subset of queries to be used later in testing the model, to
attempt to observe the same operators as encountered in that
study. We
collected the query plan at each possible cardinalities (200 in all) and
looked for ``jumps,'' that is, when the cost model for an operator in the
plan exhibited discontinuity. Jumps are determined from the estimated cost
extracted from each plan operator. We provide an example of a jump below. If
a jump is observed, we classify that particular plan operator as
discontinuous.

@@

The Exhaustive experiment gathers the cost model for each operator in each
plan for each cardinality (in our case, for cardinalities ranging from 10K
to 2M in steps of 10K, or 200 points).  Figure~\ref{fig:discontinuity}
presents for a single query\shorten{({\em query4}, exhaustive-10q-1,
  oracle)}, the {\em cost} of each of the hash-join operators utilized in
each plan at a particular cardinality. Each distinct point type depicts an
individual hash-join operator. \todo{Young: I don't understand what this
  sentence adds'' Rick: Yes, please delete the sentence.}Note that when two identical query plans
appear at two different cardinalities, we consider all the plan operators
found at the same position within these two plans to be identical. This
figure shows only a small portion of the 200 cardinalities, and is typical
(we generally observed jumps at low cardinalities for these particular queries and
relation sizes).  As shown by this figure, the hash-join operator
represented by `+' has two discontinuous jumps, both appearing below
cardinality 150K. The other hash-join operator represented by $\times$ has one
discontinuous jump.  Between the jumps, the behavior is linear. It is our
guess that
such a jump is caused by the transition between one pass of a disk-based hashing
technique to two passes (and indeed for one of the operators, perhaps the one
lower in the operator tree, the transition to three passes, hence, the two
jumps). That said, all
that matters for our predictive model is that operators that experience
such discontinuity in their cost models present opportunities for
suboptimality.

@@

\vspace{1em}
As we will see in Section~\ref{sec:experiments}, we observed only 3,347
violations (0.74\%) of non-strict monotonicity, for the largest experiment
(Experiment 7: Confirmatory) across all
the \hbox{DBMSes} we studied, justifying our conclusion that the \hbox{DBMSes} under
study are indeed monotonic.

@@

No run violated the experiment-wide sanity checks. For the largest
experiment, the Confirmatory experiment 7 discussed below, approximately
10.5\% of the query executions and 5.1\%
of the queries-at-cardinality (Q@Cs) were dropped due to query execution
and Q@C sanity checks. As a result, excessive variation in calculated query
time was observed in only 0.003\% of the Q@Cs. A total of 0.43\% of Q@C adjacent
pairs violated relaxed monotonicity and 0.74\% of the Q@Cs violated strict
monotonicity, which is acceptable.

@@

Our experiment data set consists of relational tables. There are two types of
tables. The first is a ``fixed table'' that, once created and
populated, will never be modified in the future.  In contrast, the second
type is a ``variable table.'' We alter the cardinality\c2j{ of
  this table}{, physically, of such tables} as the experiments are being performed. 
\c2j{We used a}{We organized three configurations for the data sets. The first configuration
is a small} data set with four tables, each with four integer-typed
attributes. Each table is populated with one million rows, and thus the size
of each table is roughly 16Mbyte. We also produced a version of the data set with primary
keys (of the first attribute) for all tables.

@@

\begin{table}[t]
\tbl{Experiments 1--7: Detailed Run Statistics\label{tab:run_stat}}
{%
\resizebox{140mm}{!}
{
\begin{tabular}{c|c|c|c|c|c|c|c}
& {\em Experiment}& {\em Protocol} & {\em Cumulative} & {\em Number of}&{\em
    Number} &{\em Number}& {\em Number of}\\
& & & {\em Hours} & {\em Query Instances}&{\em of Q@Cs} &{\em of QEs}&{\em Retained QEs}\\
\hline
1 & Monotonicity 		& --- & 38 & 60 & 12,000 & 12,000 & 12,000\\
2 & Refined Monotonicity 		& TTPv1 & 1,672 & 160 & 32,000 & 320,000 & 244,787\\
3 & Exhaustive 			& TTPv2 & 1,544 & 160 & 32,000 & 320,000 & 319,980\\
4 & Exhaustive with Keys 	& --- & 28 & 200 & 40,000 & --- & ---\\
5 & Initial Exploratory 	& TTPv1 & 560 & 780 & 8,842 & 88,420 & 68,891\\%40512+26367+2012
6 & Exploratory 		& TTPv2 & 1,663 & 1,200 & 12,560 & 125,600 & 114,377\\
7 & Confirmatory 		& TTPv2 & 11,375 & 7,640 & 99,558 & 995,580 & 890,631\\
\multicolumn{3}{c|}{\em Total}	& 16,880 & 10,200 & 236,960 & 1,861,600 & 1,650,666\\
\end{tabular}
}
}
\end{table}

Table~\ref{tab:run_stat} exhibits the statistics of running a number of 
queries in our experiments. All but the last column list the number of query
instances, etc., gathered by the protocol for each experiment. The last
column gives the number of query executions retained by the protocol. One of
the primary goals of TTPv2 was to reduce the number of QEs discarded due to
phantom processes that were indirectly detected. Experiments 3, 6, and 7
benefited from that protocol.

We performed six separate experiments, each looking at a different
aspect. It is important to emphasize that while the {\em queries} all came
from the same query pool and while the data sets were also shared by the
experiments, the {\em query executions} for the five experiments are
disjoint. As a side comment, we mention that for all four \hbox{DBMSes}, the query
plans generated by a \hbox{DBMS} for a particular \QatC\ of a particular query
varied between the experiments, but not between the query executions (QEs) of that query
instance at that cardinality, by virtue of the way we designed the
measurement protocol.

@@

\doubleblind{After we developed our six-step TTPv1 protocol,}{When TTPv1
    became available,} we performed our
second experiment, termed ``Refined Monotonicity,'' which more accurately tested the
(strict) monotonicity assumption, as also described in
Section~\ref{sec:suboptimality}.\todo{Young: I don't see any such discussion
  in that section..., Rick: Please take a look at the definition of the strict monotonicity assumption. We discuss this there.}
This experiment involved 160 query instances, 32,000 \QatC s, and thus
320,000 QEs, requiring 1,672 cumulative hours.  The (very small\todo{Young:
  how much?, Rick: 1.55\%}) percentage % (16312+8215)/(995321+587546) = 1.55%
of strict monotonicity violations observed was consistent with the remaining
variance of the query time measurement, concluding that none of
the \hbox{DBMSes} violated monotonicity. This also provides a validation of
our definition of suboptimality, which requires monotonicity.

We later reran this experiment using TTPv2 (we term this simply
``Exhaustive''). Note that the number of retained QEs went up a lot (that is
the main benefit of TTPv2), while the number of hours required actually
went down (another benefit).

@@

We also ran an experiment (termed ``Exhaustive with Keys'') on the Exhaustive query set but using the data
set with primary keys,  However, in this
case we did not actually execute 
the queries, but rather just examined the plans that were returned from the
\hbox{DBMS} to identify additional discontinuous operators. \comment{Thus this third
experiment was very fast.}\todo{Sabah:were any additional discontinous
  operators found? Rick: I don't know about this. I redirect this question to Sabah.}

\todo{Rick: Shouldn't this experiment  be before Experiment 3 (Exhaustive),
  because it used TTPv1?}The fifth experiment was for initial exploratory analysis of a prior
version of the causal model. (Note that this experiment used
the first version of the Tucson Timing Protocol (TTPv1)~\cite{TTPv1}.)
The earlier model had fewer independent
variables yet also had more complex interactions. Specifically, the model
did not include the independent query complexity variables of presence of
secondary indexes, presence of subquery, nor the schema complexity
independent variable of presence of secondary indexes, nor the plan space
complexity independent variable of number of repeats. The prior model also
had presence of primary keys in the schema complexity construct rather than
the query complexity construct. In reformulating that independent variable,
we were able to remove a complex mediating moderator between the plan
space complexity and suboptimality constructs.

%% The following shorten corresponds to TTPv1.
The fifth experiment was an ``Initial Exploratory'' analysis of the causal model, run on
a representative sample of queries and data sets: (a)~600 queries of one
\hbox{DBMS}, (b)~120 queries of another \hbox{DBMS}, (c)~10 queries
from each other two \hbox{DBMSes}, all without primary keys defined, and (d)~10 queries
from each of the four \hbox{DBMSes} on the primary key data set, for exploration
across all combinations, a total of 780 query instances. These query sets
are enumerated in more detail in Appendix~\ref{sec:app}.
We ran the \hbox{DBMSes} on the {\em
  change points}: adjacent cardinalities (10K apart; 300 for one \hbox{DBMS}) with different
plans. The protocol retained about 78\% of the QEs.

%% The following corresponds to TTPv2.
We later reran the exploratory analysis using TTPv2, this time on a larger sample of queries and data sets:
(a)~200 queries of four \hbox{DBMSes} without primary keys defined and
(b)~100 queries from each of the four \hbox{DBMSes} on the primary key data set,
for exploration across all combinations, a total of 1,200 query instances (= 300 $\times$ 4 (DBMSes)).
We termed this sixth experiment ``Exploratory.''  We ran the DBMSes on the {\em
  change points}: adjacent cardinalities (10K apart; 300 for one DBMS) with different
plans. This experiment required 1,663 hours. The experimental
methodology retaining 12,100 {\QatC}s (3.7\% were dropped), covering 1,123 query
instances (6.4\% were dropped). Only 443 strict monotonicity violations (0.78\%) and 301 relaxed
ones (0.54\%) were observed. This exploratory analysis allowed us to refine
the operationalizations.

The seventh and final experiment was used for confirmatory analysis of the model and thus was
called ``Confirmatory.'' This was the most time-consuming of
the experiments. Here we
used (a)~800 queries on the data set without primary keys defined and (b)~510 of
those queries that had joins on the primary key attributes, on the data set
with primary keys, 
(c)~100 queries on the data set without skew and primary keys,  
(d)~100 queries with a subquery on the data set without primary keys,  
(e)~100 queries on the data set with primary keys and secondary indexes defined, 
(f)~100 queries with a subquery on the data set with primary keys defined,  
(e)~100 queries with a subquery on the data set with primary keys and secondary indexes defined, and 
(g)~100 queries on the data set with primary keys and secondary indexes defined, 
for a total of 1,910 queries and a total of 7,640 query instances (over the
four \hbox{DBMSes}).
Again, we ran the \hbox{DBMSes} at the 99,558 {\QatC}s that were observed,
roughly 13 per query. The protocol accurately timed
both sides of adjacent {\QatC}s (53,547 change points in all).

\todo{Young: these percentages  are different from those in Section 6.1. Why? Rick: A good catch. Those numbers are out-of-date: ones from the TODS metrology paper. I replaced them with the current numbers mentioned in this para. Now consistent.}In this Confirmatory Experiment, we observed 3,347 strict monotonicity
violations (0.74\%), and 1,966 relaxed monotonicity violations (0.43\%) (out
of a total of 452,684 pairs
of two Q@Cs having identical plans and query instance), \todo{Rick:merge
  somehow with Section 6.1:}which provides
further confidence that monotonicity also applies to operators found in
queries over data in various contexts (as described above) and that our
operationalization of suboptimality is a valid one.

@@

Our goal in this paper has been to understand cost-based query optimizers
as a {\em general} class of computational artifacts and to articulate and test
a predictive model \hbox{characterizing} how such optimizers,
again, as a general class, behave.  This model can be used to further
improve \hbox{DBMSes} through engineering \hbox{efforts} that benefit from the fundamental
understanding that the scientific perspective can provide.

>>>>

%Metadata Information
\acmVolume{1}
\acmNumber{1}
\acmArticle{1}
\acmYear{2017}
\acmMonth{3}

\usepackage{url}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{graphics}
% \usepackage[total={6.5in,9in}, top=1.25in,left=0.9in]{geometry}
%\usepackage[text={6.5in,9in},centerpage,includefoot]{geometry}
% \usepackage{setspace}
\usepackage{color}
\usepackage{url}
\usepackage{balance}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{subfigure}
\usepackage{nicefrac}

@@

\begin{abstract}
The query optimization phase within a \hbox{DBMS} ostensibly finds the fastest query
execution plan from a potentially large set of enumerated plans, all of
which correctly compute the specified query. Occasionally the optimizer
selects the wrong plan, for a variety of reasons. {\em Suboptimality} is
indicated by the existence of a query plan that performs more efficiently
than the DBMS's chosen plan, for the same query. From an engineering
perspective, it is of critical importance to understand the prevalence
of suboptimality and its causal factors.
We propose a novel structural causal model to explicate the
relationship between various factors in query optimization and
suboptimality. Our model associates suboptimality with the factors of
complexity of the query and optimizer and concomitant unanticipated
interactions among the components of the optimizer. This model induces a number of specific
hypothesis that were subsequently tested on multiple \hbox{DBMSes}. We
observe that suboptimality prevalence
correlates positively with the number of operators available in the \hbox{DBMS}
(one proxy for optimizer complexity) and with the number of plans generated
by the optimizer (another proxy for optimizer complexity) and with
the number of correlation names, number of aggregates, and presence of a subquery (all measures of query complexity), providing
empirical support for this model (in fact, these causal factors explain over
half of the variance of suboptimality) as well as implications for fundamental
improvements of these optimizers. We articulate for the first time a
possible upper bound on the number of operators a DBMS may be able to support.
This paper thus provides
a new methodology to study mature query optimizers, that of empirical
generalization, proposes a novel causal model for query suboptimality, and
demonstrates that an upper bound on the number of operators is probable and
may have already been reached by one or more extant DBMSes.
\end{abstract}

@@

\item We test the six quite specific hypotheses that arise from that
  model across a wide range of DBMSes, queries, and data, showing
  through correlational and regression analysis that four
  hypothesized interactions are strongly supported, one is weakly in the
  other direction, and
  the sixth is partially supported, lending credibility to our particular model.

@@

\begin{figure}[h!]\centering
\includegraphics[width=0.60\textwidth]{figures/org_ncpq.pdf}
\caption{Cumulative percentage of queries \hbox{exhibiting} the \hbox{indicated}
  number of plan changes\label{fig:planchanges}}
\end{figure}

@@

This is why we needed a new methodology. We want to understand cost-based
optimization deeply. This means that we need to go beyond the examination of
a single query optimizer, as is done in almost every paper (with a few
important exceptions~\cite{harish07,Haritsa10,Leis15}) over the forty-year history of
query optimization research, to study multiple instances of that
optimization architecture, in an effort to achieve generalizable results.

@@

Early investigation of plan suboptimality resulted in approaches such as
dynamic query-reoptimization~\cite{Avnur,Bellamkonda13,kabra98,Li07}, which exploit more
accurate runtime statistics that appear while a query is being executed, to
steer in-flight plan reoptimization. The very presence of such a radical
change to the normal optimize-execute sequence indicates that plan
suboptimality was of interest to some researchers.

@@

\vspace{0.6em}
We feel that skewed data (in the ``data complexity''
construct) will not impact plan space complexity. Rather, our expectation is
that it will negatively impact accuracy of plan cost estimation, and thus
increase suboptimality directly.

@@

\vspace{0.6em}
The schema complexity construct consists of the variable ``Presence of
secondary indexes.'' Such indexes provide opportunities for the optimizer to
consider more candidate plans that use these indexes, due to the additional
query evaluation operators now applicable. Those additional plans enable the optimizer to
possibly do a better job, while also adding complexity to the optimization
process.  The quality of query optimization ``very much depends on how much
the query engine relies on [cardinality] estimates and on how complex the
physical database design is, i.e., the number of indexes
available.''~\cite{Leis15}.

We hypothesize that this factor has a more complex role in the
model, serving as a {\em moderator} of two interactions previously
introduced. We hypothesize that the overall effect of the secondary indexes
is to increase the strength of the interaction between query
complexity to plan space complexity. (Note that as the presence of primary
key is required for secondary indexes, we don't include that former
independent variable here.)

@@

We define the skew as ``the reciprocal of the number of distinct
values,'' so $0 < \hbox{\em skew} \leq 1$.
For 2M distinct values, the skew would be $\nicefrac{1}{2M}= 0.000005$, which is practically 0. 
For exactly one distinct value, the skew would be 1.0. For two distinct values, 
the skew would be 0.5. For ten distinct values, the skew would be 0.1.

@@

We use two values of skew: $\nicefrac{1}{2M}$ (termed {\em tiny}) and
$\nicefrac{1}{10K}$ (termed {\em small}).
\shorten{Let's first illustrate with a skew factor of 1.0. 
This translates to exactly 1 value for the entire table, or 2M spans.
To shorten, we remove 10K spans, equal to 10K values. 
This removal keeps the skew at 1.0, due to the unique values in the
remaining spans.
At the end, 10K spans, each with a single value, remain.

@@

At the other end of the spectrum, consider a skew factor of
1/2M.  This skew factor, which is almost 0, gets translated to 2M
distinct values.  So}For the former, we generate a single span of 2M values.  To shorten the
table, it makes no sense to remove this single span in its entirety. But we
can remove 10K values from this single span.  Note that this changes the
skew to 1/1.99M, then eventually to 1/10K, which is still very close to zero
(the skew has changed from .0000005 to .00001).
For the latter, we generate 2,000 spans each of 10K tuples, and drop a span
to reduce to 1.99M tuples, repeating.
\shorten{We use five skewness values: 0.000005 (which we term only slightly
misleadingly as ``no skew''), 0.001, 0.1, 0.5, and 1.0,
for this independent variable.}

@@

To classify each operator as continuous or discontinuous, we use the data
from the ``Exhaustive'' experiments to be described in detail in Section~\ref{sec:experiments}. We
used a small subset of queries to be used later in testing the model, to
attempt to observe the same operators as encountered in that
study. We
collected the query plan at each possible cardinalities (200 in all) and
looked for ``jumps,'' that is, when the cost model for an operator in the
plan exhibited discontinuity. Jumps are determined from the estimated cost
extracted from each plan operator. We provide an example of a jump below. If
a jump is observed, we classify that particular plan operator as
discontinuous.

@@

The Exhaustive experiments gather the cost model for each operator in each
plan for each cardinality (in our case, for cardinalities ranging from 10K
to 2M in steps of 10K, or 200 points).  Figure~\ref{fig:discontinuity}
presents for a single query\shorten{({\em query4}, exhaustive-10q-1,
  oracle)}, the {\em cost} of each of the hash-join operators utilized in
each plan at a particular cardinality. Each distinct point type depicts an
individual hash-join operator.  This
figure shows only a small portion of the 200 cardinalities, and is typical
(we generally observed jumps at low cardinalities for these particular queries and
relation sizes).  As shown by this figure, the hash-join operator
represented by `+' has two discontinuous jumps, both appearing below
cardinality 150K. The other hash-join operator represented by $\times$ has one
discontinuous jump.  Between the jumps, the behavior is linear. It is our
guess that
such a jump is caused by the transition between one pass of a disk-based hashing
technique to two passes (and indeed for one of the operators, perhaps the one
lower in the operator tree, the transition to three passes, hence, the two
jumps). That said, all
that matters for our predictive model is that operators that experience
such discontinuity in their cost models present opportunities for
suboptimality.

@@

\vspace{1em}
That said, the suspicious monotonicity violations raised the fundamental question of whether
we were using a sufficiently reliable timing method. The answer was {\em no}, as
the experiment used a (popular but) naive measurement technique 
based on end-to-end timing of which the results are 
typically perturbed by system noise potentially incurred during the measurement. 
We realized that the accuracy of this timing data was
insufficient for both this monotonicity test but also for our suboptimality study. 
We then switched to a new protocol, TTPv1~\cite{TTPv1}, and later TTPv2~\cite{TTPv2},
which fortunately is able to obtain quite accurate and precise timing.

Specifically, as we will see in Section~\ref{sec:experiments}, 
using the protocol we observed only 3,347 violations (0.74\%) of non-strict monotonicity, 
for the largest experiment (Experiment 7: Confirmatory) across all
the \hbox{DBMSes} we studied. 
Hence, this result justified our conclusion that the \hbox{DBMSes} under
study are indeed monotonic.

@@

No run violated the experiment-wide sanity checks. For the largest
experiment, the Confirmatory Experiment (number 7) discussed below, approximately
10.5\% of the query executions and 5.1\%
of the queries-at-cardinality (Q@Cs) were dropped due to query execution
and Q@C sanity checks. As a result, excessive variation in calculated query
time was observed in only 0.003\% of the Q@Cs. A total of 0.43\% of Q@C adjacent
pairs violated relaxed monotonicity and 0.74\% of the Q@Cs violated strict
monotonicity, which is acceptable.

@@

Our experiment data set consists of relational tables. There are two types of
tables. The first is a ``fixed table'' that, once created and
populated, will never be modified in the future.  In contrast, the second
type is a ``variable table.'' We alter the cardinality\c2j{ of
  this table}{, physically, of such tables} as the experiments are being performed. 
\c2j{We used a}{We generated four configurations for the data sets. The first configuration
is a small} data set with four tables, each with four integer-typed
attributes. Each table is populated with one million rows, and thus the size
of each table is roughly 16Mbyte, with tiny skew (cf.~Section~\ref{sec:datacomplexity}). We also produced versions of the data set
with (i)~small skew, (ii)~primary keys (of the first attribute), and (iii)~primary
keys and secondary indexes, for all tables, as detailed in Appendix~\ref{sec:appdatasets}.

@@

\begin{table}[t]
\tbl{Experiments 1--7: Detailed Run Statistics\label{tab:run_stat}}
{%
\resizebox{140mm}{!}
{
\begin{tabular}{c|c|c|c|c|c|c|c}
& {\em Experiment}& {\em Protocol} & {\em Cumulative} & {\em Number of}&{\em
    Number} &{\em Number}& {\em Number of}\\
& & & {\em Hours} & {\em Query Instances}&{\em of Q@Cs} &{\em of QEs}&{\em Retained QEs}\\
\hline
1 & Monotonicity 		& --- & 38 & 60 & 12,000 & 12,000 & 12,000\\
2 & Exhaustive		& TTPv1 & 1,672 & 160 & 32,000 & 320,000 & 244,787\\
3\shorten{4} & Exhaustive with Keys 	& --- & 28 & 200 & 40,000 & --- & ---\\
4\shorten{5} & Initial Exploratory 	& TTPv1 & 560 & 780 & 8,842 & 88,420 & 68,891\\%40512+26367+2012
5\shorten{3} & Refined Exhaustive 	& TTPv2 & 1,544 & 160 & 32,000 & 320,000 & 319,980\\
6 & Exploratory 		& TTPv2 & 1,663 & 1,200 & 12,560 & 125,600 & 114,377\\
7 & Confirmatory 		& TTPv2 & 11,375 & 7,640 & 99,558 & 995,580 & 890,631\\
\multicolumn{3}{c|}{\em Total}	& 16,880 & 10,200 & 236,960 & 1,861,600 & 1,650,666\\
\end{tabular}
}
}
\end{table}

We performed seven separate experiments, each looking at a different
aspect. \hbox{Table~\ref{tab:run_stat}} exhibits the statistics of running a number of 
queries in our experiments. All but the last column list the number of query
instances, etc., gathered by the protocol for each experiment. The last
column gives the number of query executions retained by the protocol. One of
the primary goals of TTPv2 was (i)~to reduce the number of QEs discarded due to
phantom processes that were indirectly detected and (ii)~to collect relevant measures 
directly associated with I/O, such as BlockIO Delay time~\cite{TTPv2}. 
Experiments 5, 6, and~7 benefited from that protocol. Other details about
the experiments can be found in Appendix~\ref{sec:otherdetails}.

The queries came from 19 query sets, summarized below with more details in
\hbox{Appendix~\ref{sec:querysets}}. It is important to emphasize that while the {\em queries} all came
from the same query pool and while the data sets were also shared by the
experiments (see the Appendix for details), the {\em query executions} for
the six experiments, except Experiment~3 with no timing and hence no QEs, 
are disjoint. As a side comment, we mention that for all four \hbox{DBMSes}, the query
plans generated by a \hbox{DBMS} for a particular \QatC\ of a particular query
varied between the experiments, but not between the query executions (QEs)
of that Q@C, by virtue of the way \doubleblind{we designed the
measurement protocol}{the measurement protocol was designed}.

@@

\doubleblind{After we developed our six-step TTPv1 protocol,}{When TTPv1
    became available,} we performed our
second experiment, termed ``Exhaustive,'' which more accurately tested the 
monotonicity assumption. 
This experiment involved 160 query instances, 32,000 \QatC s (thus the name:
we timed each query at {\em all} cardinalities, generating 2,000 \QatC s for
each query instance), and thus
320,000 QEs (ten QEs for each \QatC ), requiring 1,672 cumulative hours.  The (very small: 1.6\%)
percentage % (16312+8215)/(995321+587546) = 1.6%
of strict monotonicity violations observed was consistent with the remaining
variance of the query time measurement, concluding that none of
the \hbox{DBMSes} violated monotonicity. This also provides a validation of
our definition of suboptimality, which requires monotonicity.

@@

We also ran an experiment (termed ``Exhaustive with Keys'') on the 
Exhaustive query set but using the data set with primary keys,  However, in this
case we did not actually execute
the queries, but rather just examined the plans that were returned from the
\hbox{DBMS} to identify additional discontinuous operators. (Thus this third
experiment was very fast.)\todo{Sabah:were any additional discontinuous
  operators found? Rick: Yes, I think they were. like {\tt eq\_ref} on MySQL. But for more details 
I redirect this question to Sabah.}

The fourth experiment was an ``Initial Exploratory'' 
analysis of a prior version of the causal model.
That earlier model had fewer independent
variables yet also had more complex interactions. Specifically, the model
did not include the independent query complexity variables of presence of
secondary indexes, presence of subquery, nor the schema complexity
independent variable of presence of secondary indexes, nor the plan space
complexity independent variable of number of repeats. The prior model also
had presence of primary keys in the schema complexity construct rather than
the query complexity construct. In reformulating that independent variable,
we were able to remove a complex mediating moderator between the plan
space complexity and suboptimality constructs.

%% The following shorten corresponds to TTPv1.
In the experiment we ran a representative sample of queries and data sets: (a)~600 queries of one
\hbox{DBMS}, (b)~120 queries of another \hbox{DBMS}, (c)~10 queries
from each other two \hbox{DBMSes}, all without primary keys defined, and (d)~10 queries
from each of the four \hbox{DBMSes} on the primary key data set, for exploration
across all combinations, a total of 780 query instances.  (Again, consult
Appendix~\ref{sec:querysets} for details on these query sets.)
We ran the \hbox{DBMSes} only on the {\em
  change points}: adjacent cardinalities (10K apart; 300 for one \hbox{DBMS}) with different
plans (unlike the Exhaustive experiments). The protocol retained about 78\% of the QEs.

As a large number of QEs (roughly a quarter) were discarded by TTPv1 in
these experiments, and as these discarded QEs were correlated with query
time (and thus perhaps with suboptimality),
\doubleblind{we sought to develop a better measurement protocol, which was
  the impetus for TTPv2~\cite{TTPv2}}{we switched to a better measurement
  protocol that had recently become available: TTPv2~\cite{TTPv2}}.
%Our subsequent experiments indeed benefited from the improved protocol, producing 
%highly lossless ($\sim$5\% Q@C drop) but very clean ($\sim$1\% violation) data.
The fifth experiment was to rerun the Exhaustive experiment  (we term this simply
``Refined Exhaustive'') using that protocol. The main purpose of this experiment 
was to reexamine monotonicity within in the enhanced
protocol.\todo{Sabah:didn't we also check discontinuity once more? If so,
  were any new discontinuous operators found in this third Exhaustive experiment?}

We observed that the monotonicity violation rate decreased
from 1.54\% to 1.02\% for strict monotonicity and from 0.61\% to 0.60\% for
non-strict monotonicity, thereby reaffirming our monotonicity assumption
(the first benefit). Moreover, the number of retained QEs went up
significantly (compare244K to 320K, a second and the primary benefit of TTPv2), 
while the number of hours required actually went down somewhat (a third benefit).
These benefits reassured us of the quality of data 
for subsequent experiments via the improved protocol.

%% The following corresponds to TTPv2.
We later reran the exploratory analysis using TTPv2, this time on a larger sample of queries and data sets:
(a)~200 queries of four \hbox{DBMSes} without primary keys defined and
(b)~100 queries from each of the four \hbox{DBMSes} on the primary key data set,
for exploration across all combinations, a total of 1,200 query instances. % (= 300 $\times$ 4 (DBMSes))
We termed this sixth experiment ``Exploratory.'' This experiment 
retained 12,100 {\QatC}s (3.7\% were dropped), concerning 1,123 query
instances (6.4\% were dropped). Only 443 strict monotonicity violations (0.78\%) and 301 relaxed
violation (0.54\%) were observed. This exploratory analysis allowed us to refine
the operationalizations.

The seventh and final experiment was used for confirmatory analysis of the model and thus was
called ``Confirmatory.'' This was the most time-consuming of
the experiments. Here we
used (a)~800 queries on the data set without primary keys defined and (b)~510 of
those queries that had joins on the primary key attributes, on the data set
with primary keys, 
(c)~100 queries on the data set with tiny skew and primary keys,  
(d)~100 queries with a subquery on the data set without primary keys,  
(e)~100 queries on the data set with primary keys and secondary indexes defined, 
(f)~100 queries with a subquery on the data set with primary keys defined,  
(e)~100 queries with a subquery on the data set with primary keys and secondary indexes defined, and 
(g)~100 queries on the data set with primary keys and secondary indexes defined, 
for a total of 1,910 queries and a total of 7,640 query instances (over the
four \hbox{DBMSes}). (Again, see Appendix~\ref{sec:querysets} for details.)
We ran the \hbox{DBMSes} at the 99,558 {\QatC}s that were observed,
roughly 13 per query. The protocol accurately timed
both sides of adjacent {\QatC}s (53,547 change points in all).

In this Confirmatory Experiment, we observed 3,347 strict monotonicity
violations (0.74\%), and 1,966 relaxed monotonicity violations (0.43\%) (out
of a total of 452,684 pairs
of two Q@Cs having identical plans and query instance), which provides
further confidence that monotonicity also applies to operators found in
queries over data in various contexts (as described above) and that our
operationalization of suboptimality is a valid one.

@@

Our goal in this paper has been to understand cost-based query optimizers
as a \hbox{\em general} class of computational artifacts and to articulate and test
a predictive model \hbox{characterizing} how such optimizers,
again, as a general class, behave.  This model can be used to further
improve \hbox{DBMSes} through engineering \hbox{efforts} that benefit from the fundamental
understanding that the scientific perspective can provide.
